---
title: "Working with Missing Reasons in the Tidyverse"
author: "Kyle Husmann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Background

In many datasets, reasons for missing values are interlaced with data as codes
or strings. For example:

```{r}
na_codes_csv <- "
person_id,age,favorite_color
1,20,BLUE
1,-99,BLUE
2,21,RED
3,30,-98
4,31,-99
5,41,RED
6,50,-98
"

na_strings_csv <- "
person_id,age,favorite_color
1,20,BLUE
1,_DECLINED_ANSWER_,BLUE
2,21,RED
3,30,_TECHNICAL_ERROR_
4,31,_DECLINED_ANSWER_
5,41,RED
6,50,_TECHNICAL_ERROR_
"
```

Presently, it is easy to convert these values to NA when loading via
`read_delim` et al.:

```{r}
read_csv(na_codes_csv, na=c("-99", "-98"))

read_csv(na_strings_csv, na=c("_DECLINED_ANSWER_", "_TECHNICAL_ERROR_"))
```

Challenges arise, however, when we want to utilize these missing reasons while 
data wrangling. For example, consider the following things we might want to do:

1. Create a sample that only includes rows where individuals declined to answer
at least one item.

2. Find the distribution of reasons for individuals missing the "favorite_color"
item.

3. Find the average age for individuals that had technical errors reporting
their favorite color.

4. Find the average age for individuals that are missing a report of their
favorite color, *for each missing reason*

Because we're wanting to work with missing reasons, the `na` option in 
`read_csv` isn't useful anymore... instead, we need to load our CSVs
our CSVs as strings, perform manipulations, mask out the missing reasons
(if necessary), and then type_convert back to values:

```{r}
string_df <- read_csv(na_strings_csv, col_types="c")

na_values <- c("_DECLINED_ANSWER_", "_TECHNICAL_ERROR_")

col_types <- cols(
  person_id = col_integer(),
  age = col_number(),
  favorite_color = col_factor(levels=c("RED", "BLUE")),
)

# 1. Create a sample that only includes rows where individuals declined to asnwer
# items.

string_df %>%
  filter(
    if_any(everything(), \(x) x == "_DECLINED_ANSWER_")
  ) %>%
  type_convert(
    col_types=col_types,
    na=na_values,
  )

# 2. Find the distribution of reasons for individuals missing the "favorite_color"
# item.

string_df %>%
  mutate(
    across(
      everything(),
      \(x) if_else(x %in% na_values, x, NA_character_)
    )
  ) %>%
  count(favorite_color)

# 3. Find the average age for individuals that had technical errors reporting
# their favorite color.

string_df %>%
  filter(
    favorite_color == "_TECHNICAL_ERROR_",
  ) %>%
  type_convert(
    col_types=col_types,
    na=na_values,
  ) %>%
  summarize(
    age = mean(age, na.rm=TRUE)
  )

# 4. Find the average age for individuals that are missing a report of their
# favorite color, *for each missing reason*

string_df %>%
  mutate(
    favorite_color = if_else(
      favorite_color %in% na_values,
      favorite_color,
      NA_character_,
    ),
    age = parse_number(age, na = na_values),
  ) %>%
  summarize(
    age = mean(age, na.rm=T),
    .by=favorite_color,
  )
```

Although the above code works, it's not pretty. Because our missing reasons
are a different type from our columns, we have to keep making string comparisons
and value casts in order to get the result we want. It means we have to keep the 
`na_values` and `col_types` vars hanging around, and apply them at the right
times and in the right orders to accomplish our desired tasks.

## An alternative approach

An alternative approach is to put values and missing reasons into separate 
columns:

```{r}
values_df <- string_df %>%
  type_convert(
    col_types=col_types,
    na=na_values,
  )

missing_df <- string_df %>%
  mutate(
    across(
      everything(),
      \(x) if_else(x %in% na_values, x, NA_character_)
    )
  ) %>%
  rename_with(\(x) str_c(x, "_missing"))

better_df <- bind_cols(values_df, missing_df)

better_df
```

This allows us to do all of the value and missing reasons parsing a single time,
and then use the structure of the data to accomplish all the same tasks as
before... without having to constantly check for missing reasons and cast value
types:

```{r}
# 1. Create a sample that only includes rows where individuals declined to answer
# at least one item.

# Compare to:
# string_df %>%
#   filter(
#     if_any(everything(), \(x) x == "_DECLINED_ANSWER_")
#   ) %>%
#   type_convert(
#     col_types=col_types,
#     na=na_values,
#   )

better_df %>%
  filter(
    if_any(ends_with("_missing"), \(x) x == "_DECLINED_ANSWER_")
  )


# 2. Find the distribution of reasons for individuals missing the "favorite_color"
# item.

# Compare to:
# string_df %>%
#   mutate(
#     across(
#       everything(),
#       \(x) if_else(x %in% na_values, x, NA_character_)
#     )
#   ) %>%
#   count(favorite_color)

better_df %>%
  count(favorite_color_missing)

# 3. Find the average age for individuals that had technical errors reporting
# their favorite color.

# Compare to:
# string_df %>%
#   filter(
#     favorite_color == "_TECHNICAL_ERROR_",
#   ) %>%
#   type_convert(
#     col_types=col_types,
#     na=na_values,
#   ) %>%
#   summarize(
#     age = mean(age, na.rm=TRUE)
#   )

better_df %>%
  filter(favorite_color_missing == "_TECHNICAL_ERROR_") %>%
  summarize(
    age = mean(age, na.rm=T)
  )

# 4. Find the average age for individuals that are missing a report of their
# favorite color, *for each missing reason*

# Compare to:
# string_df %>%
#   mutate(
#     favorite_color = if_else(
#       favorite_color %in% na_values,
#       favorite_color,
#       NA_character_,
#     ),
#     age = parse_number(age, na = na_values),
#   ) %>%
#   summarize(
#     age = mean(age, na.rm=T),
#     .by=favorite_color,
#   )

better_df %>%
  summarize(
    age = mean(age, na.rm=T),
    .by = favorite_color_missing,
  )
```

The resulting code is MUCH easier to read and no longer littered with type
conversions and missingness checks. We also don't need to carry `na_values`
and `col_types` around with us anywhere, because the value columns have their
type information in the dataframe (as they should!), and we know if we're
working with a missing reason if we're in a *_missing column. (In effect, it
enables us to interact with dataframe values as tagged union types)

With this simple example, the difference may seem small, but as operations & data
become more complex, this approach saves many headaches!

## `readr` feature request

Given how common it is for formats to interlace values and missing reasons, I 
think it would be worth giving `readr` the option to load missing reasons in
addition to values. I propose adding a `channels` argument (or something 
similar) to `read_delim` et al. that would give users a flexible API around 
loading values alongside missing reasons. Here's a rough example of how it might
look:

(Note this is just a rough, example implementation to illustrate the idea,
not optimized at all!)

```{r}
read_delim_ext <- function(file, delim, na = c("", "NA"), col_types = NULL, 
                           channels = "values", ...) {
  default_suffixes <- c(
    values = "_values",
    missing = "_missing"
  )
  
  if (is.null(names(channels))) {
    names(channels) <- rep_along(channels, "")
  }
 
  if (length(channels) == 1 && names(channels) == "") {
    channel_suffixes <- ""
  } else {
    channel_suffixes <- map2_vec(
      channels,
      names(channels),
      \(x, n) if (n == "") { default_suffixes[[x]] } else { x }
    )
  } 
 
  channel_names <- map2_vec(
    channels,
    names(channels),
    \(x, n) if_else(n == "", x, n)
  )
  
  channels <- set_names(channel_suffixes, channel_names)
  
  string_df <- read_delim(
    file,
    delim,
    col_types = cols(.default=col_character()),
    ...
  )
  
  result <- list()
  
  if ("values" %in% channel_names) {
    values_df <- string_df %>%
       type_convert(
         col_types=col_types,
         na=na,
       ) %>%
       rename_with(\(x) str_c(x, channels[["values"]]))
    result <- append(result, values_df)
  }
  
  if ("missing" %in% channel_names) {
    missing_df <- string_df %>%
      mutate(
        across(
          everything(),
          \(x) if_else(x %in% na, x, NA_character_)
        )
      ) %>%
      rename_with(\(x) str_c(x, channels[["missing"]]))
    result <- append(result, missing_df)
  }  
  bind_cols(result)
}
```

Here's how it would be used:

```{r}
# Load values only (default behavior, same as before):

read_delim_ext(
  na_strings_csv,
  delim=",",
  na=na_values,
  col_types=col_types
  # channels = "values" <- implicit default
)

# Load missing reasons:

read_delim_ext(
  na_strings_csv,
  delim=",",
  na=na_values,
  col_types=col_types,
  channels="missing",
)

# Load values AND missing reasons:

read_delim_ext(
  na_strings_csv,
  delim=",",
  na=na_values,
  col_types=col_types,
  channels=c("values", "missing"),
)

# Load values and missing reasons, but without a suffix on values:

read_delim_ext(
  na_strings_csv,
  delim=",",
  na=na_values,
  col_types=col_types,
  channels=c(values="", "missing"),
)


# Load values and missing reasons with custom column suffixes:

read_delim_ext(
  na_strings_csv,
  delim=",",
  na=na_values,
  col_types=col_types,
  channels=c(values="_v", missing="_m"),
)
```

As you can see, when loading a single channel (`values` or `missing`), no
suffixes are appended to column names. When both channels are loaded, columns
get suffixes that can be customized by making the `channels` arg a named vector.

This is just a first pass at this idea -- I'd appreciate other thoughts &
feedback!
