---
title: "Labelled Columns with Frictionless-R"
author: "Kyle Husmann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(jsonlite)

datapackage_txt <- '
{
  "name": "value-label-examples",
  "resources": [
    {
      "name": "example1",
      "path": "example1.csv",
      "profile": "tabular-data-resource",
      "schema": {
        "fields": [
          {
            "name": "day_scale",
            "description": "On a scale from 1-5, how was your day?",
            "type": "integer",
            "constraints": {
              "enum": [1, 2, 3, 4, 5]
            },
            "enumOrdered": true,
            "enumLabels": {
              "1": "Very bad",
              "2": "Bad",
              "3": "Neutral",
              "4": "Good",
              "5": "Very good"
            }
          },
          {
            "name": "pain_scale",
            "description": "On a scale from 0-9, how do you rate your pain?",
            "type": "integer",
            "constraints": {
              "enum": [0, 1, 2, 3, 4, 5, 6, 7, 8, 9]
            },
            "enumOrdered": true,
            "enumLabels": {
              "0": "No pain",
              "9": "Worst pain imaginable"
            }
          },
          {
            "name": "pick_a_number",
            "description": "Pick a number!",
            "type": "integer",
            "enumLabels": {
              "7": "Magic number"
            }
          }         
        ]
      }
    }
  ]
}
'

cat(datapackage_txt, file="example-pkg/datapackage.json")

set.seed(9001)

tibble(
  day_scale = round(runif(10, 1, 5)),
  pain_scale = round(runif(10, 0, 9)),
  pick_a_number = c(runif(4, 1, 20), 7, runif(5, 1, 20))
) %>%
  write_csv("example-pkg/example1.csv")

```

## Introduction

I've put together this vignette to demo a proof of concept implementation of
value labels for `frictionless-r`. My goal has been a simple,
ergonomic API for working with value (and missing) labels in frictionless
that generalizes the concepts introduced by SAS, Stata, SPSS, et al., and can 
be used with tidyverse wrangling functions in powerfully
expressive ways. Luckily there was a lot of existing work in this area I was able
to build on!

NOTE THAT THIS IMPLEMENTATION IS HIGHLY EXPERIMENTAL! There's a lot to still
implement (e.g. ordered factors). Even more, I'm still learning all
the quirks of R's class system and how `tidyverse` augments it. This is a proof-
of-concept designed to explore these ideas and get some early feedback on these
directions that I'm exploring.

For all the examples in this text, I will use the example datapackage loaded
below (and [linked here](example-pkg/datapackage.json)):

```{r}
library(tidyverse)
library(frictionless)
library(jsonlite)
library(haven)

pkg <- read_package("example-pkg/datapackage.json")
```

## Background

Value labels are extra metadata
attached to certain values of a variable in a dataset. For example, a question
on a survey asking an individual to rate how good their day was on a scale from
1-5 might include labels alongside the number to help "calibrate" the numeric
responses for the respondent (e.g. 1 = Very bad, 2 = Bad, 3 = Neutral, 4 = Good,
5 = Very good). Sometimes these labels are referred to as "anchors" for the scale,
especially when they label the scale's extrema (e.g. a 10-point pain scale with
0 = No pain, and 9 = Worst pain imaginable).

In the [recently accepted frictionless extension](https://github.com/frictionlessdata/specs/pull/844),
proposed by Phil Schumm, value labels can be added to any variable type via the
`enumLabels` schema field property. The `enumOrdered` property is was
also introduced to distinguish ordinal vs categorical variables. With these
properties, the aforementioned "good day scale" and "pain scale" can be
represented as follows:

```{r}
get_schema(pkg, "example1")$fields[[1]] %>%
  toJSON(pretty=TRUE, auto_unbox=TRUE)


get_schema(pkg, "example1")$fields[[2]] %>%
  toJSON(pretty=TRUE, auto_unbox=TRUE)
```

Although the `enumLabels` property has `enum` in its name, it is not limited to
types with enum constraints -- it can be used to provide labels for any type.
for example, the following field definition is also valid:

```{r}
get_schema(pkg, "example1")$fields[[3]] %>%
  toJSON(pretty=TRUE, auto_unbox=TRUE)
```

(By contrast, the `enumOrdered` only applies to fields with `enum` constraints.
I think this is a little confusing, and might be worth revisiting in the
standard extension at some point.)

(Note that my implementation here does not yet support `enumOrdered`, but it
should be easy to add.)

## Frictionless-r implementation

Here's what it looks like when the above three fields are loaded into a
dataframe with my frictionless-r implementation (found in my fork on the
[`categorical-ext` branch](https://github.com/khusmann/frictionless-r/tree/categorical-ext)):

```{r}
(example1 <- read_resource(pkg, "example1"))
```

As you can see, the `tibble` displays both values and their labels. You can't 
see it in the rendered Rmd, but in the R console, the labels are a grayed out
color to help differentiate them from the actual values.

This behavior accomplished by way of the `haven` package in the `tidyverse` universe.
`haven` provides the `haven_labelled` class that enables numeric and character
vectors to be augmented by labels. A limitation of the `haven_labelled` class,
however, is that it can only be used with numeric and character vectors,
because it was originally designed as a compatibility layer for the variable
types found in SPSS, Stata, and SAS.

This behavior is fine for loading numeric fields with labels in frictionless
schemas, like `pick_a_number`. For labeled enum types, like `day_scale` and 
`pain_scale` however, conversion to labelled numerics does not preserve the
`enum` constraint in its type. As you can see above, to address this, I extended
the `haven_labelled` class into a new `frictionless_labelled_fct` class that
also support factors as a base type (denoted by the `fct+lbl` tibble header).

The advantage of this approach is that when we remove the labels using 
`haven::zap_labels`, we get factor types for these enum fields, instead of
raw numeric:

```{r}
zap_labels(example1)
```

This means that all the levels of the enum described in the schema are properly
present in the factor levels, even if they don't all show up in the data:

```{r}
levels(zap_labels(example1)$day_scale)
```

Had we used only a labelled numeric, removing the labels would have given us a
raw numeric type that we'd have to turn into a factor... and would have required
another visit to the field's schema to know what the levels were supposed to be.
Instead, by loading all this information up front into the custom type,
manipulations become much more fluid.

## Working with value labels

`haven` provides `as_factor` for converting values to labels. By default, it
will keep the original values around as levels:

```{r}
as_factor(example1)
```

Alternatively, unlabeled values can be marked NA with the `levels="labels`
option:

```{r}
as_factor(example1, levels="labels")
```

Between `zap_labels()` and `as_factor()` working with values or labels becomes a breeze:

```{r}
example1 %>%
  filter(pain_scale == "9") # Note -- this is a string comparison for now, but I plan to fix later

example1 %>%
  filter(as_factor(pain_scale) == "Worst pain imaginable")
```

More examples:

```{r}
example1 %>%
  group_by(day_scale) %>%
  summarize(
    avg_pain = mean(as.numeric(as.character(pain_scale))) # Note -- this is awkward for now (see above comment about strings)
  )
```

This is especially nice for graphing. (Note the x axis labels):

```{r}
example1 %>%
  ggplot(aes(day_scale)) +
  geom_histogram(stat="count")

example1 %>%
  ggplot(aes(as_factor(day_scale))) +
  geom_histogram(stat="count")
```

The downside of `as_factor` and `zap_labels` is they lose the label attribute
information. I've been experimenting with an alternative approach that allows
for round-trip conversions between labels and values. For now, I'm calling
the function `apply_labels()`:

```{r}
apply_labels(example1)

# Round trip
apply_labels(apply_labels(example1))
```

## Variable labels

As an added bonus, `haven_labelled` also stores "variable labels", a string
of metadata to hold information about fields. We can have `frictionless-r` load
metadata into this attribute as well. In the present implementation, I have it
grab data from the `description` field property, so this information is
displayed when viewing the table with `View()`.

Other packages like [`codebookr`](https://brad-cannell.github.io/codebookr/) and
[`labelled`](https://cran.r-project.org/web/packages/labelled/vignettes/intro_labelled.html)
use this info to generate codebooks:

```{r}
library(labelled)
library(gt)
generate_dictionary(example1) %>%
  gt()
```

I'm working on my own interactive codebook generator as well, stay tuned!

## Missing values

Value labels in SAS, Stata, SPSS sometimes add metadata to missing values as
well (e.g. missing reasons). Although the `haven` package provides additional
functionality for working with labelled missingness, its approach is very much
tied to the specific quriks of SAS, Stata, and SPSS'
implementations.

I have found that simply loading missingness as separate columns provides an
implementation-agnostic way of loading and working with labelled missingness.
I describe this [approach in a separate vignette](missing_reasons.html).

This approach for working with missing reasons is highly synergistic with the
value labels I've described in this document. It gives us a simple general
interface by which we can work with SAS, SPSS, and Stata formatted data in R,
and easily convert between their styles of missingness representations. I will
talk about this more in a future vignette.

Thanks for reading this far! I appreciate your thoughts and feedback.
